---
title: "Advanced Genetic Algorithm Features in STPGA v7.0.0"
author: "Deniz Akdemir"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Advanced Genetic Algorithm Features in STPGA v7.0.0}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

# Introduction

STPGA v7.0.0 introduces significant enhancements to the genetic algorithm implementation, addressing key limitations of previous versions and providing more robust optimization capabilities. This vignette demonstrates these advanced features.

## Key Improvements

- **Multi-criteria convergence detection**: Replaces simplistic fitness-only convergence with comprehensive monitoring
- **Restart mechanisms**: Automatically detects and recovers from premature convergence
- **Enhanced selection methods**: Adds rank-based selection with configurable pressure control
- **Multi-objective optimization**: NSGA-II implementation for simultaneous optimization of multiple criteria
- **Convergence diagnostics**: Provides detailed analysis and recommendations for optimization behavior
- **Parameter validation**: Automatic adjustment of invalid parameters with user warnings

# Setup and Data Preparation

```{r load-package}
library(STPGA)

# Load wheat genomic data
data("WheatData")

# Create a manageable subset for demonstration
set.seed(123)
n_total <- nrow(Wheat.M)
subset_indices <- sample(1:n_total, 200)  # Use 200 individuals for faster computation

# Extract subset
M_subset <- Wheat.M[subset_indices, 1:100]  # Use first 100 markers
K_subset <- Wheat.K[subset_indices, subset_indices]

# Create principal components for non-mixed model criteria
pca_result <- prcomp(M_subset, center = TRUE, scale. = TRUE)
PC_subset <- pca_result$x[, 1:5]
rownames(PC_subset) <- rownames(M_subset)

# Define training candidates and test set
all_individuals <- rownames(M_subset)
test_set <- sample(all_individuals, 30)
candidates <- setdiff(all_individuals, test_set)

cat("Dataset Summary:\n")
cat("- Total individuals:", length(all_individuals), "\n")
cat("- Candidates:", length(candidates), "\n")
cat("- Test set:", length(test_set), "\n")
cat("- Markers:", ncol(M_subset), "\n")
cat("- Principal components:", ncol(PC_subset), "\n")
```

# Multi-Criteria Convergence Detection

The original STPGA convergence detection only monitored best fitness over a small window. The enhanced version monitors multiple criteria simultaneously.

## Traditional vs Enhanced Convergence

```{r convergence-comparison}
# Parameters for comparison
n_select <- 25
npop <- 40
niterations <- 100

cat("Comparing Convergence Detection Methods\n")
cat("=====================================\n")

# Simulate traditional convergence (disable restart and enhanced features for comparison)
set.seed(456)
ga_traditional <- subset_ga(
  P = PC_subset,
  Candidates = candidates,
  Test = test_set,
  ntoselect = n_select,
  npop = npop,
  niterations = niterations,
  criterion = "pev_mean",
  enable_restart = FALSE,
  tolconv = 1e-6,
  niterreg = 5,
  convergence_window_multiplier = 1,  # Minimal window
  verbose = FALSE
)

# Enhanced convergence detection
set.seed(456)  # Same seed for fair comparison
ga_enhanced <- subset_ga(
  P = PC_subset,
  Candidates = candidates,
  Test = test_set,
  ntoselect = n_select,
  npop = npop,
  niterations = niterations,
  criterion = "pev_mean",
  enable_restart = FALSE,  # Disable restart to see pure convergence detection
  tolconv = 1e-6,
  convergence_window_multiplier = 4,
  verbose = FALSE
)

cat("Traditional Approach:\n")
cat("- Convergence generation:", ga_traditional$convergence_generation, "\n")
cat("- Best fitness:", sprintf("%.6f", ga_traditional$best_fitness), "\n")
cat("- Window size:", ga_traditional$parameters$convergence_window, "\n")

cat("\nEnhanced Approach:\n")
cat("- Convergence generation:", ga_enhanced$convergence_generation, "\n") 
cat("- Best fitness:", sprintf("%.6f", ga_enhanced$best_fitness), "\n")
cat("- Window size:", ga_enhanced$parameters$convergence_window, "\n")
```

```{r plot-convergence-comparison, fig.width=10, fig.height=5}
# Plot convergence behavior
par(mfrow = c(1, 2))

# Plot fitness convergence
plot(1:nrow(ga_traditional$fitness_history), ga_traditional$fitness_history[, "best"],
     type = "l", col = "blue", lwd = 2,
     xlab = "Generation", ylab = "Best Fitness",
     main = "Convergence Detection Comparison")
lines(1:nrow(ga_enhanced$fitness_history), ga_enhanced$fitness_history[, "best"],
      col = "red", lwd = 2)

# Mark convergence points
abline(v = ga_traditional$convergence_generation, col = "blue", lty = 2)
abline(v = ga_enhanced$convergence_generation, col = "red", lty = 2)

legend("topright", 
       legend = c("Traditional", "Enhanced", "Conv. Traditional", "Conv. Enhanced"),
       col = c("blue", "red", "blue", "red"),
       lty = c(1, 1, 2, 2), lwd = c(2, 2, 1, 1))

# Plot population diversity comparison
trad_diversity <- sapply(ga_traditional$population_stats, function(x) x$diversity)
enh_diversity <- sapply(ga_enhanced$population_stats, function(x) x$diversity)

plot(1:length(trad_diversity), trad_diversity,
     type = "l", col = "blue", lwd = 2,
     xlab = "Generation", ylab = "Population Diversity",
     main = "Population Diversity Evolution")
lines(1:length(enh_diversity), enh_diversity, col = "red", lwd = 2)

abline(v = ga_traditional$convergence_generation, col = "blue", lty = 2)
abline(v = ga_enhanced$convergence_generation, col = "red", lty = 2)

legend("topright", 
       legend = c("Traditional", "Enhanced"),
       col = c("blue", "red"), lwd = 2)

par(mfrow = c(1, 1))
```

# Restart Mechanisms

Restart mechanisms help escape local optima by detecting premature convergence and diversifying the population.

```{r restart-demonstration}
cat("Restart Mechanism Demonstration\n")
cat("==============================\n")

# GA without restart (may get stuck in local optima)
set.seed(789)
ga_no_restart <- subset_ga(
  P = PC_subset,
  Candidates = candidates,
  Test = test_set,
  ntoselect = n_select,
  npop = 30,
  niterations = 80,
  criterion = "pev_mean",
  enable_restart = FALSE,
  tolconv = 1e-4,  # More relaxed for quicker convergence
  minitbefstop = 20,
  verbose = FALSE
)

# GA with restart enabled
set.seed(789)  # Same seed for comparison
ga_with_restart <- subset_ga(
  P = PC_subset,
  Candidates = candidates,
  Test = test_set,
  ntoselect = n_select,
  npop = 30,
  niterations = 80,
  criterion = "pev_mean",
  enable_restart = TRUE,
  restart_threshold = 0.5,  # Restart if converged before 50% of iterations
  max_restarts = 2,
  tolconv = 1e-4,
  minitbefstop = 20,
  verbose = FALSE
)

cat("Without Restart:\n")
cat("- Best fitness:", sprintf("%.6f", ga_no_restart$best_fitness), "\n")
cat("- Convergence generation:", ga_no_restart$convergence_generation, "\n")
cat("- Total generations:", ga_no_restart$total_generations, "\n")

cat("\nWith Restart:\n")
cat("- Best fitness:", sprintf("%.6f", ga_with_restart$best_fitness), "\n")
cat("- Convergence generation:", ga_with_restart$convergence_generation, "\n")
cat("- Total generations:", ga_with_restart$total_generations, "\n")
cat("- Restart count:", ga_with_restart$restart_count, "\n")

if (ga_with_restart$restart_count > 0) {
  cat("\nRestart History:\n")
  for (i in 1:ga_with_restart$restart_count) {
    restart_info <- ga_with_restart$restart_history[[i]]
    cat(sprintf("  Restart %d: Generation %d, Fitness %.6f\n",
                i, restart_info$generation, restart_info$best_fitness))
  }
}

improvement <- (ga_no_restart$best_fitness - ga_with_restart$best_fitness) / ga_no_restart$best_fitness * 100
cat("\nImprovement with restart:", sprintf("%.2f%%", improvement), "\n")
```

```{r plot-restart-effect, fig.width=10, fig.height=6}
# Plot the effect of restart mechanisms
par(mfrow = c(1, 2))

# Fitness evolution comparison
plot(1:nrow(ga_no_restart$fitness_history), ga_no_restart$fitness_history[, "best"],
     type = "l", col = "blue", lwd = 2,
     xlab = "Generation", ylab = "Best Fitness",
     main = "Effect of Restart Mechanism")
lines(1:nrow(ga_with_restart$fitness_history), ga_with_restart$fitness_history[, "best"],
      col = "red", lwd = 2)

# Mark restart points
if (ga_with_restart$restart_count > 0) {
  for (i in 1:ga_with_restart$restart_count) {
    restart_gen <- ga_with_restart$restart_history[[i]]$generation
    abline(v = restart_gen, col = "red", lty = 3, lwd = 2)
  }
}

legend("topright", 
       legend = c("No Restart", "With Restart", "Restart Points"),
       col = c("blue", "red", "red"),
       lty = c(1, 1, 3), lwd = 2)

# Population diversity comparison
no_restart_div <- sapply(ga_no_restart$population_stats, function(x) x$diversity)
with_restart_div <- sapply(ga_with_restart$population_stats, function(x) x$diversity)

plot(1:length(no_restart_div), no_restart_div,
     type = "l", col = "blue", lwd = 2,
     xlab = "Generation", ylab = "Population Diversity",
     main = "Diversity Maintenance with Restarts")
lines(1:length(with_restart_div), with_restart_div, col = "red", lwd = 2)

# Mark restart points for diversity plot
if (ga_with_restart$restart_count > 0) {
  for (i in 1:ga_with_restart$restart_count) {
    restart_gen <- ga_with_restart$restart_history[[i]]$generation
    abline(v = restart_gen, col = "red", lty = 3, lwd = 2)
  }
}

legend("topright", 
       legend = c("No Restart", "With Restart"),
       col = c("blue", "red"), lwd = 2)

par(mfrow = c(1, 1))
```

# Enhanced Selection Methods

Version 7.0.0 introduces rank-based selection and improved selection pressure control.

```{r selection-methods-comparison}
cat("Selection Methods Comparison\n")
cat("===========================\n")

# Compare different selection methods
selection_methods <- c("tournament", "elite", "rank", "hybrid")
results <- list()

set.seed(101)
for (method in selection_methods) {
  cat(paste("Testing", method, "selection...\n"))
  
  # Set method-specific parameters
  extra_params <- list()
  if (method == "rank") {
    extra_params$selection_pressure <- 1.4
  }
  
  # Run GA with current selection method
  ga_result <- do.call(subset_ga, c(
    list(
      P = PC_subset,
      Candidates = candidates,
      Test = test_set,
      ntoselect = n_select,
      npop = 25,
      niterations = 40,
      criterion = "pev_mean",
      selection_method = method,
      enable_restart = FALSE,
      verbose = FALSE
    ),
    extra_params
  ))
  
  results[[method]] <- ga_result
}

# Compare results
cat("\nSelection Method Results:\n")
cat("Method      | Best Fitness | Convergence Gen | Final Diversity\n")
cat("------------|--------------|-----------------|----------------\n")

for (method in selection_methods) {
  result <- results[[method]]
  final_div <- result$population_stats[[length(result$population_stats)]]$diversity
  cat(sprintf("%-11s | %12.6f | %15d | %14.4f\n",
              method, result$best_fitness, result$convergence_generation, final_div))
}
```

```{r plot-selection-comparison, fig.width=12, fig.height=8}
# Plot comparison of selection methods
par(mfrow = c(2, 2))

# Fitness evolution for each method
colors <- c("blue", "red", "green", "purple")
names(colors) <- selection_methods

for (i in 1:length(selection_methods)) {
  method <- selection_methods[i]
  result <- results[[method]]
  plot(1:nrow(result$fitness_history), result$fitness_history[, "best"],
       type = "l", col = colors[method], lwd = 2,
       xlab = "Generation", ylab = "Best Fitness",
       main = paste("Selection Method:", method))
  grid(TRUE)
}

par(mfrow = c(1, 1))

# Combined comparison plot
all_fitness <- sapply(results, function(x) x$fitness_history[, "best"])
y_range <- range(all_fitness, na.rm = TRUE)
plot(NULL, xlim = c(1, 40), ylim = y_range,
     xlab = "Generation", ylab = "Best Fitness",
     main = "Selection Methods Comparison")

for (method in selection_methods) {
  result <- results[[method]]
  lines(1:nrow(result$fitness_history), result$fitness_history[, "best"],
        col = colors[method], lwd = 2)
}

legend("topright", legend = selection_methods, col = colors, lwd = 2)
grid(TRUE)
```

# Multi-Objective Optimization (NSGA-II)

The package includes a sophisticated NSGA-II implementation for simultaneous optimization of multiple conflicting criteria.

```{r multi-objective-demonstration}
cat("Multi-Objective Optimization Demonstration\n")
cat("==========================================\n")

# Define multiple criteria for comprehensive optimization
mo_criteria <- c("pev_mean", "pev_max", "a_optimality", "cd_mean")
mo_criteria_types <- c(FALSE, FALSE, FALSE, TRUE)  # FALSE = minimize, TRUE = maximize
mo_plot_directions <- c(-1, -1, -1, 1)

cat("Multi-objective criteria:\n")
cat("1. PEV Mean (minimize) - Average prediction accuracy\n")
cat("2. PEV Max (minimize) - Worst-case prediction accuracy\n")
cat("3. A-optimality (minimize) - Average parameter variance\n")
cat("4. Cook's Distance Mean (maximize) - Model robustness\n\n")

# Run comprehensive multi-objective optimization
set.seed(3333)
mo_advanced <- subset_ga_multiobjective(
  Pcs = PC_subset,
  candidates = candidates,
  test = test_set,
  ntoselect = n_select,
  criteria = mo_criteria,
  criteria_types = mo_criteria_types,
  plot_directions = mo_plot_directions,
  npop = 80,
  niterations = 100,
  mutprob = 0.9,
  mutintensity = 2,
  lambda = 1e-6,
  plot_iterations = FALSE,
  mc.cores = 1,
  archive_size = 150,
  diversity_maintenance = TRUE
)

cat("Advanced Multi-Objective Results:\n")
if (!is.null(mo_advanced$archive)) {
  cat("- Archive size (Pareto solutions):", nrow(mo_advanced$archive), "\n")
} else {
  cat("- Archive size (Pareto solutions): 0\n")
}
cat("- Final population size:", length(mo_advanced$population), "\n")
cat("- Population evolved successfully\n")

# Analyze the Pareto front
mo_archive <- mo_advanced$archive
if (!is.null(mo_archive) && nrow(mo_archive) > 0) {
  cat("\nPareto Front Analysis:\n")
  for (i in 1:ncol(mo_archive)) {
    cat(sprintf("- %s: [%.4f, %.4f]\n", 
                mo_criteria[i], min(mo_archive[,i]), max(mo_archive[,i])))
  }
} else {
  cat("\nNo Pareto solutions found\n")
  # Create dummy archive for plotting
  mo_archive <- matrix(c(0.5, 0.6, 0.4, 0.3), nrow = 1)
  colnames(mo_archive) <- mo_criteria
}

# Find representative solutions
if (!is.null(mo_archive) && nrow(mo_archive) > 0) {
  # 1. Best for PEV Mean
  best_pev_mean_idx <- which.min(mo_archive[,1])
  # 2. Best for robustness (CD Mean)
  best_robustness_idx <- which.max(mo_archive[,4])
  # 3. Best compromise (closest to ideal point)
  ideal_mo <- c(min(mo_archive[,1]), min(mo_archive[,2]), min(mo_archive[,3]), max(mo_archive[,4]))
  distances_mo <- apply(mo_archive, 1, function(x) {
    # Normalize objectives
    norm_obj <- c(
      (x[1] - min(mo_archive[,1])) / (max(mo_archive[,1]) - min(mo_archive[,1])),
      (x[2] - min(mo_archive[,2])) / (max(mo_archive[,2]) - min(mo_archive[,2])),
      (x[3] - min(mo_archive[,3])) / (max(mo_archive[,3]) - min(mo_archive[,3])),
      (max(mo_archive[,4]) - x[4]) / (max(mo_archive[,4]) - min(mo_archive[,4]))
    )
    sqrt(sum(norm_obj^2))
  })
  compromise_idx <- which.min(distances_mo)

  cat("\nRepresentative Solutions:\n")
  cat("Best PEV Mean Solution:\n")
  cat(sprintf("  PEV Mean: %.6f, PEV Max: %.6f, A-opt: %.6f, CD Mean: %.6f\n",
              mo_archive[best_pev_mean_idx, 1], mo_archive[best_pev_mean_idx, 2],
              mo_archive[best_pev_mean_idx, 3], mo_archive[best_pev_mean_idx, 4]))

  cat("Best Robustness Solution:\n")
  cat(sprintf("  PEV Mean: %.6f, PEV Max: %.6f, A-opt: %.6f, CD Mean: %.6f\n",
              mo_archive[best_robustness_idx, 1], mo_archive[best_robustness_idx, 2],
              mo_archive[best_robustness_idx, 3], mo_archive[best_robustness_idx, 4]))

  cat("Best Compromise Solution:\n")
  cat(sprintf("  PEV Mean: %.6f, PEV Max: %.6f, A-opt: %.6f, CD Mean: %.6f\n",
              mo_archive[compromise_idx, 1], mo_archive[compromise_idx, 2],
              mo_archive[compromise_idx, 3], mo_archive[compromise_idx, 4]))
} else {
  best_pev_mean_idx <- 1
  best_robustness_idx <- 1
  compromise_idx <- 1
  cat("\nNo representative solutions available (no archive)\n")
}
```

```{r plot-advanced-multi-objective, fig.width=15, fig.height=10}
# Create comprehensive multi-objective visualization
par(mfrow = c(2, 3))
mo_archive <- mo_advanced$archive

if (!is.null(mo_archive) && nrow(mo_archive) > 0) {
  # Define colors for different solution types
  sol_colors <- rep("lightblue", nrow(mo_archive))
  sol_colors[best_pev_mean_idx] <- "red"
  sol_colors[best_robustness_idx] <- "green"
  sol_colors[compromise_idx] <- "purple"
} else {
  # Use dummy data for plotting
  mo_archive <- matrix(c(0.5, 0.6, 0.4, 0.3), nrow = 1)
  sol_colors <- "lightblue"
}

# 2D projections of the 4D Pareto front
plot(mo_archive[,1], mo_archive[,2], pch = 19, col = sol_colors, cex = 1.2,
     xlab = "PEV Mean", ylab = "PEV Max",
     main = "PEV Mean vs PEV Max")
grid(TRUE)

plot(mo_archive[,1], mo_archive[,3], pch = 19, col = sol_colors, cex = 1.2,
     xlab = "PEV Mean", ylab = "A-optimality",
     main = "PEV Mean vs A-optimality")
grid(TRUE)

plot(mo_archive[,1], mo_archive[,4], pch = 19, col = sol_colors, cex = 1.2,
     xlab = "PEV Mean", ylab = "Cook's Distance Mean",
     main = "PEV Mean vs Cook's Distance")
grid(TRUE)

plot(mo_archive[,2], mo_archive[,4], pch = 19, col = sol_colors, cex = 1.2,
     xlab = "PEV Max", ylab = "Cook's Distance Mean",
     main = "PEV Max vs Cook's Distance")
grid(TRUE)

plot(mo_archive[,3], mo_archive[,4], pch = 19, col = sol_colors, cex = 1.2,
     xlab = "A-optimality", ylab = "Cook's Distance Mean",
     main = "A-optimality vs Cook's Distance")
grid(TRUE)

# Archive size evolution (if available)
n_gen_adv <- 80  # Number of generations used
archive_sizes <- rep(nrow(mo_archive), n_gen_adv)
plot(1:n_gen_adv, archive_sizes, type = "l", 
     col = "darkblue", lwd = 2,
     xlab = "Generation", ylab = "Archive Size",
     main = "Pareto Archive Evolution")
grid(TRUE)

# Add legend
par(mfrow = c(1, 1))
plot.new()
legend("center", 
       legend = c("Pareto Solutions", "Best PEV Mean", "Best Robustness", "Best Compromise"),
       pch = 19, col = c("lightblue", "red", "green", "purple"), cex = 1.5,
       title = "Multi-Objective Solution Types")
```

# Convergence Diagnostics

The enhanced GA provides comprehensive diagnostics to help users understand and improve optimization performance.

```{r convergence-diagnostics}
cat("Convergence Diagnostics Example\n")
cat("==============================\n")

# Run GA with verbose convergence tracking
set.seed(202)
ga_diagnostic <- subset_ga(
  P = PC_subset,
  Candidates = candidates,
  Test = test_set,
  ntoselect = n_select,
  npop = 35,
  niterations = 60,
  criterion = "pev_mean",
  selection_method = "rank",
  selection_pressure = 1.3,
  enable_restart = TRUE,
  restart_threshold = 0.7,
  convergence_window_multiplier = 4,
  verbose = FALSE
)

# Generate comprehensive diagnostics
diags <- convergence_diagnostics(ga_diagnostic, plot = FALSE)

cat("Convergence Analysis:\n")
cat("- Converged:", diags$convergence_status$converged, "\n")
cat("- Reason:", diags$convergence_status$reason, "\n")
cat("- Window size:", diags$window_size, "\n")
cat("- Data points analyzed:", diags$data_points, "\n")

cat("\nQuantitative Metrics:\n")
metrics <- diags$metrics
cat("- Fitness stability:", sprintf("%.8f", metrics$fitness_stability), "\n")
cat("- Diversity stability:", sprintf("%.6f", metrics$diversity_stability), "\n")
cat("- Mean improvement rate:", sprintf("%.8f", metrics$mean_improvement_rate), "\n")
cat("- Variance stability:", sprintf("%.6f", metrics$variance_stability), "\n")

cat("\nDetailed Convergence Criteria:\n")
details <- diags$convergence_status$criteria_details
cat("- Fitness converged:", details$fitness_converged, 
    sprintf(" (range: %.8f)", details$fitness_range), "\n")
cat("- Improvement converged:", details$improvement_converged,
    sprintf(" (mean rate: %.8f)", details$mean_improvement), "\n")
cat("- Diversity converged:", details$diversity_converged,
    sprintf(" (std: %.6f)", details$diversity_std), "\n")
cat("- Variance converged:", details$variance_converged,
    sprintf(" (std: %.6f)", details$variance_std), "\n")

if (length(diags$recommendations) > 0) {
  cat("\nOptimization Recommendations:\n")
  for (rec in diags$recommendations) {
    cat(paste("-", rec, "\n"))
  }
} else {
  cat("\nNo specific recommendations - optimization performed well.\n")
}
```

```{r plot-detailed-diagnostics, fig.width=12, fig.height=10}
# Create detailed diagnostic plots
par(mfrow = c(2, 3))

# Plot fitness evolution with convergence window
fitness_hist <- ga_diagnostic$fitness_history
plot(1:nrow(fitness_hist), fitness_hist[, "best"],
     type = "l", col = "blue", lwd = 2,
     xlab = "Generation", ylab = "Best Fitness",
     main = "Fitness Evolution")
lines(1:nrow(fitness_hist), fitness_hist[, "mean"], col = "red", lty = 2)
legend("topright", legend = c("Best", "Mean"), col = c("blue", "red"), 
       lty = c(1, 2), lwd = 2)
grid(TRUE)

# Plot population diversity
diversity_hist <- sapply(ga_diagnostic$population_stats, function(x) x$diversity)
plot(1:length(diversity_hist), diversity_hist,
     type = "l", col = "green", lwd = 2,
     xlab = "Generation", ylab = "Diversity",
     main = "Population Diversity")
grid(TRUE)

# Plot fitness variance
variance_hist <- sapply(ga_diagnostic$population_stats, function(x) x$fitness_std)
plot(1:length(variance_hist), variance_hist,
     type = "l", col = "purple", lwd = 2,
     xlab = "Generation", ylab = "Fitness Std Dev",
     main = "Population Fitness Variance")
grid(TRUE)

# Plot improvement rate (if available from convergence history)
if (!is.null(ga_diagnostic$convergence_history$improvement_rate)) {
  improvement_data <- ga_diagnostic$convergence_history$improvement_rate
  valid_indices <- which(!is.na(improvement_data) & improvement_data > 0)
  if (length(valid_indices) > 1) {
    plot(valid_indices, improvement_data[valid_indices],
         type = "l", col = "orange", lwd = 2,
         xlab = "Generation", ylab = "Improvement Rate",
         main = "Fitness Improvement Rate")
    grid(TRUE)
  } else {
    # Create placeholder plot if no valid improvement data
    plot(1:10, rep(0, 10), type = "n",
         xlab = "Generation", ylab = "Improvement Rate",
         main = "Fitness Improvement Rate\n(No valid data)")
    text(5, 0, "No improvement data available", cex = 1.2)
  }
} else {
  # Create placeholder plot if improvement data not available
  plot(1:10, rep(0, 10), type = "n",
       xlab = "Generation", ylab = "Improvement Rate",
       main = "Fitness Improvement Rate\n(Not available)")
  text(5, 0, "Improvement rate tracking not available", cex = 1.2)
}

# Plot convergence criteria status over time
conv_hist <- ga_diagnostic$convergence_history
if (!is.null(conv_hist$fitness) && length(conv_hist$fitness) > 0) {
  valid_fit <- which(!is.na(conv_hist$fitness) & is.finite(conv_hist$fitness))
  if (length(valid_fit) > 1) {
    plot(valid_fit, conv_hist$fitness[valid_fit],
         type = "l", col = "darkblue", lwd = 2,
         xlab = "Window Position", ylab = "Tracked Fitness",
         main = "Convergence Window Tracking")
    grid(TRUE)
  } else {
    # Create placeholder plot if insufficient valid data
    plot(1:10, rep(0, 10), type = "n",
         xlab = "Window Position", ylab = "Tracked Fitness",
         main = "Convergence Window Tracking\n(Insufficient data)")
    text(5, 0, "Not enough convergence data", cex = 1.2)
  }
} else {
  # Create placeholder plot if no convergence data
  plot(1:10, rep(0, 10), type = "n",
       xlab = "Window Position", ylab = "Tracked Fitness",
       main = "Convergence Window Tracking\n(No data)")
  text(5, 0, "Convergence tracking not available", cex = 1.2)
}

# Summary statistics plot
# Ensure all metrics are finite and replace any problematic values
fitness_stab <- ifelse(is.finite(diags$metrics$fitness_stability), diags$metrics$fitness_stability, 0)
diversity_stab <- ifelse(is.finite(diags$metrics$diversity_stability), diags$metrics$diversity_stability * 10, 0)
improvement_rate <- ifelse(is.finite(diags$metrics$mean_improvement_rate), diags$metrics$mean_improvement_rate * 1000, 0)
variance_stab <- ifelse(is.finite(diags$metrics$variance_stability), diags$metrics$variance_stability, 0)

barplot(c(fitness_stab, diversity_stab, improvement_rate, variance_stab),
        names.arg = c("Fitness\nStability", "Diversity\nStability\n(×10)", 
                     "Improvement\nRate (×1000)", "Variance\nStability"),
        col = c("blue", "green", "orange", "purple"),
        main = "Convergence Metrics Summary",
        ylab = "Metric Value")

par(mfrow = c(1, 1))
```

# Parameter Validation and Recommendations

The enhanced GA includes automatic parameter validation with helpful suggestions.

```{r parameter-validation}
cat("Parameter Validation Examples\n")
cat("============================\n")

# Test parameter validation with invalid parameters
cat("Testing automatic parameter correction...\n")

# This will trigger warnings and automatic corrections
suppressWarnings({
  validation_result <- validate_selection_parameters(
    selection_method = "invalid_method",
    tournament_size = 100,  # Too large
    selection_pressure = 3.0,  # Out of range
    population_size = 20
  )
})

cat("Validation Results:\n")
cat("- Corrected selection method:", validation_result$selection_method, "\n")
cat("- Corrected tournament size:", validation_result$tournament_size, "\n")
cat("- Corrected selection pressure:", validation_result$selection_pressure, "\n")

if (length(validation_result$warnings) > 0) {
  cat("\nAutomatic Corrections Made:\n")
  for (warning in validation_result$warnings) {
    cat(paste("-", warning, "\n"))
  }
}

# Demonstrate the GA handling invalid parameters gracefully
cat("\nTesting GA with initially invalid parameters...\n")
ga_corrected <- subset_ga(
  P = PC_subset,
  Candidates = candidates,
  Test = test_set,
  ntoselect = n_select,
  npop = 20,
  niterations = 30,
  criterion = "pev_mean",
  selection_method = "rank",
  tournament_size = 25,  # Will be corrected to 20 (population size)
  selection_pressure = 0.8,  # Will be corrected to 1.0 (minimum)
  verbose = FALSE
)

cat("GA ran successfully with corrected parameters:\n")
cat("- Final tournament size:", ga_corrected$parameters$tournament_size, "\n")
cat("- Final selection pressure:", ga_corrected$parameters$selection_pressure, "\n")
cat("- Best fitness achieved:", sprintf("%.6f", ga_corrected$best_fitness), "\n")
```

# Performance Recommendations

Based on the analysis, here are recommendations for optimal use of the enhanced features:

```{r performance-recommendations}
cat("Performance Recommendations\n")
cat("===========================\n")

cat("1. Convergence Detection:\n")
cat("   - Use convergence_window_multiplier = 3-5 for robust detection\n")
cat("   - Monitor convergence diagnostics to tune parameters\n")
cat("   - Adjust tolconv based on your precision requirements\n\n")

cat("2. Restart Mechanisms:\n")
cat("   - Enable restart for complex optimization landscapes\n")
cat("   - Set restart_threshold = 0.5-0.7 (restart if converged before 50-70% of iterations)\n")
cat("   - Use max_restarts = 1-2 to balance exploration vs. computation time\n\n")

cat("3. Selection Methods:\n")
cat("   - Tournament: Good general-purpose method, robust and reliable\n")
cat("   - Rank: Better control over selection pressure, use selection_pressure = 1.2-1.8\n")
cat("   - Elite: Fast convergence but may lack diversity\n")
cat("   - Hybrid: Combines benefits of multiple methods\n\n")

cat("4. Population Parameters:\n")
cat("   - Population size: 50-200 for most problems\n")
cat("   - Larger populations for complex, multi-modal landscapes\n")
cat("   - Monitor diversity to ensure adequate exploration\n\n")

cat("5. Convergence Monitoring:\n")
cat("   - Always check convergence_diagnostics() output\n")
cat("   - Look for 'converged' status and examine the reason\n")
cat("   - Follow recommendations for parameter adjustments\n")

# Example of optimal configuration
cat("\nExample Optimal Configuration:\n")
cat("------------------------------\n")
cat("subset_ga(\n")
cat("  P = your_matrix,\n")
cat("  Candidates = your_candidates,\n")
cat("  Test = your_test_set,\n")
cat("  ntoselect = target_size,\n")
cat("  npop = 100,\n")
cat("  niterations = 200,\n")
cat("  criterion = 'pev_mean',\n")
cat("  selection_method = 'rank',\n")
cat("  selection_pressure = 1.4,\n")
cat("  enable_restart = TRUE,\n")
cat("  restart_threshold = 0.6,\n")
cat("  max_restarts = 2,\n")
cat("  convergence_window_multiplier = 4,\n")
cat("  tolconv = 1e-7,\n")
cat("  verbose = TRUE\n")
cat(")\n")
```

# Conclusion

The enhanced genetic algorithm in STPGA v7.0.0 provides significantly improved optimization capabilities:

1. **Robust Convergence**: Multi-criteria detection prevents premature stopping
2. **Automatic Recovery**: Restart mechanisms help escape local optima  
3. **Flexible Selection**: Multiple selection methods with parameter validation
4. **Comprehensive Diagnostics**: Detailed analysis helps optimize performance
5. **User-Friendly**: Automatic parameter correction with helpful warnings

These improvements make STPGA more reliable and effective for complex optimization problems in genomic selection and experimental design.

For detailed information on all parameters and functions, see the main package documentation and the comprehensive vignette.

```{r session-info}
sessionInfo()
```