---
title: "STPGA: Subset Training Population Genetic Algorithm"
author: "Deniz Akdemir"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{STPGA: Subset Training Population Genetic Algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

The **STPGA** (Subset Training Population Genetic Algorithm) package provides advanced tools for optimizing training set selection in genomic prediction and experimental design. This vignette demonstrates the package's key functionalities using real wheat genomic data.

## Key Features

- **Modern optimization criteria**: A-, D-, E-optimality, PEV, Cook's Distance
- **Mixed model support**: BLUP-based criteria for genomic selection
- **Genetic algorithms**: Single and multi-objective optimization
- **Flexible interface**: Unified criterion function with legacy compatibility
- **Robust numerics**: Ridge regularization and numerical stability

# Getting Started

Let's start by loading the package and exploring the included wheat dataset:

```{r load-package}
library(STPGA)

# Load the wheat dataset
data("WheatData")

# Explore the data structure
cat("Dataset components:\n")
cat("- Wheat.Y: Phenotype data (", nrow(Wheat.Y), " observations)\n")
cat("- Wheat.M: Marker data (", nrow(Wheat.M), " individuals x ", ncol(Wheat.M), " markers)\n")
cat("- Wheat.K: Kinship matrix (", nrow(Wheat.K), " x ", ncol(Wheat.K), ")\n")
```

```{r explore-data}
# Look at the phenotype data
head(Wheat.Y)

# Check marker data dimensions and first few markers
cat("Marker matrix preview:\n")
Wheat.M[1:5, 1:5]

# Examine kinship matrix structure
cat("Kinship matrix diagonal (first 10):\n")
diag(Wheat.K)[1:10]
cat("Kinship matrix off-diagonal (first 5x5):\n")
Wheat.K[1:5, 1:5]
```

# Optimization Criteria

STPGA provides several optimization criteria for training set selection. Let's demonstrate each using the wheat data.

## Setup for Demonstrations

```{r setup-demo}
# Define candidate pool and test sets
all_individuals <- rownames(Wheat.M)
n_total <- length(all_individuals)

# Create training candidates and test set
set.seed(123)
test_set <- sample(all_individuals, 40)
candidates <- setdiff(all_individuals, test_set)

cat("Total individuals:", n_total, "\n")
cat("Test set size:", length(test_set), "\n")
cat("Candidate pool size:", length(candidates), "\n")

# Extract first 5 principal components for non-mixed model criteria
# This reduces computational complexity for high-dimensional marker data
cat("\nExtracting first 5 principal components for non-mixed model criteria...\n")
pca_result <- prcomp(Wheat.M, center = TRUE, scale. = TRUE)
Wheat.PC5 <- pca_result$x[, 1:5]
rownames(Wheat.PC5) <- rownames(Wheat.M)

cat("PC matrix dimensions:", nrow(Wheat.PC5), "x", ncol(Wheat.PC5), "\n")
cat("Variance explained by first 5 PCs:", sprintf("%.2f%%", 100 * sum(pca_result$sdev[1:5]^2) / sum(pca_result$sdev^2)), "\n")
```

## Classical Optimality Criteria

The package implements the classical A-, D-, and E-optimality criteria from experimental design theory.

```{r classical-criteria}
# Select training sets of different sizes
train_small <- sample(candidates, 30)
train_medium <- sample(candidates, 60)
train_large <- sample(candidates, 100)

# Compute A-optimality using first 5 PCs (minimizes average prediction variance)
aopt_small <- a_optimality(train_small, test_set, Wheat.PC5)
aopt_medium <- a_optimality(train_medium, test_set, Wheat.PC5)
aopt_large <- a_optimality(train_large, test_set, Wheat.PC5)

cat("A-optimality results:\n")
cat("Small training (n=30):", sprintf("%.2e", aopt_small), "\n")
cat("Medium training (n=60):", sprintf("%.2e", aopt_medium), "\n")
cat("Large training (n=100):", sprintf("%.2e", aopt_large), "\n")
cat("Improvement (large vs small):", sprintf("%.1fx", aopt_small/aopt_large), "\n")

# D-optimality using first 5 PCs (minimizes generalized variance)
dopt_small <- d_optimality(train_small, test_set, Wheat.PC5)
dopt_medium <- d_optimality(train_medium, test_set, Wheat.PC5)
dopt_large <- d_optimality(train_large, test_set, Wheat.PC5)

cat("\nD-optimality results:\n")
cat("Small training (n=30):", sprintf("%.2f", dopt_small), "\n")
cat("Medium training (n=60):", sprintf("%.2f", dopt_medium), "\n")
cat("Large training (n=100):", sprintf("%.2f", dopt_large), "\n")

# E-optimality using first 5 PCs (minimizes maximum variance)
eopt_small <- e_optimality(train_small, test_set, Wheat.PC5)
eopt_medium <- e_optimality(train_medium, test_set, Wheat.PC5)
eopt_large <- e_optimality(train_large, test_set, Wheat.PC5)

cat("\nE-optimality results:\n")
cat("Small training (n=30):", sprintf("%.2f", eopt_small), "\n")
cat("Medium training (n=60):", sprintf("%.2f", eopt_medium), "\n")
cat("Large training (n=100):", sprintf("%.2f", eopt_large), "\n")
```

## Prediction Error Variance (PEV) Criteria

PEV criteria are particularly relevant for prediction accuracy in genomic selection.

```{r pev-criteria}
# Mean PEV using first 5 PCs
pev_mean_small <- pev_mean(train_small, test_set, Wheat.PC5)
pev_mean_large <- pev_mean(train_large, test_set, Wheat.PC5)

# Normalized PEV using first 5 PCs (scale-invariant)
pev_norm_small <- pev_mean(train_small, test_set, Wheat.PC5, normalized = TRUE)
pev_norm_large <- pev_mean(train_large, test_set, Wheat.PC5, normalized = TRUE)

# Maximum PEV using first 5 PCs (worst-case scenario)
pev_max_small <- pev_max(train_small, test_set, Wheat.PC5)
pev_max_large <- pev_max(train_large, test_set, Wheat.PC5)

cat("PEV Results:\n")
cat("Mean PEV (small):", sprintf("%.2e", pev_mean_small), "\n")
cat("Mean PEV (large):", sprintf("%.2e", pev_mean_large), "\n")
cat("Improvement:", sprintf("%.1fx", pev_mean_small/pev_mean_large), "\n")

cat("\nNormalized PEV:\n")
cat("Small training:", sprintf("%.6f", pev_norm_small), "\n")
cat("Large training:", sprintf("%.6f", pev_norm_large), "\n")

cat("\nMax PEV:\n")
cat("Small training:", sprintf("%.2e", pev_max_small), "\n")
cat("Large training:", sprintf("%.2e", pev_max_large), "\n")
cat("Max/Mean ratio (small):", sprintf("%.2f", pev_max_small/pev_mean_small), "\n")
cat("Max/Mean ratio (large):", sprintf("%.2f", pev_max_large/pev_mean_large), "\n")
```

## Coefficient of Determination (R²) Criteria

The Coefficient of Determination measures the proportion of variance explained by the model.

```{r cook-distance}
# Mean Coefficient of Determination (R²) using first 5 PCs
cd_mean_small <- cd_mean(train_small, test_set, Wheat.PC5)
cd_mean_large <- cd_mean(train_large, test_set, Wheat.PC5)

# Normalized Coefficient of Determination using first 5 PCs
cd_norm_small <- cd_mean(train_small, test_set, Wheat.PC5, normalized = TRUE)
cd_norm_large <- cd_mean(train_large, test_set, Wheat.PC5, normalized = TRUE)

cat("Coefficient of Determination Results:\n")
cat("Mean CD (small):", sprintf("%.2e", cd_mean_small), "\n")
cat("Mean CD (large):", sprintf("%.2e", cd_mean_large), "\n")

cat("\nNormalized R²:\n")
cat("Small training:", sprintf("%.6f", cd_norm_small), "\n")
cat("Large training:", sprintf("%.6f", cd_norm_large), "\n")
```

# Mixed Model Criteria for Genomic Selection

For genomic selection applications, the package provides mixed model criteria that account for genetic relationships.

```{r mixed-models}
# Select smaller sets for computational efficiency with mixed models
train_mm <- sample(candidates, 50)
test_mm <- sample(test_set, 20)

# Extract relevant kinship submatrices
K_train <- Wheat.K[train_mm, train_mm]
K_test <- Wheat.K[test_mm, test_mm]
K_test_train <- Wheat.K[test_mm, train_mm]

# Mixed model PEV with different heritability scenarios
h2_high <- list(Vg = 0.8, Ve = 0.2)  # High heritability (80%)
h2_low <- list(Vg = 0.3, Ve = 0.7)   # Low heritability (30%)

# Use smaller dataset for computational efficiency
# For mixed models, use subset of individuals but all relationships
all_mm_individuals <- c(train_mm, test_mm)
P_subset <- Wheat.M[all_mm_individuals, 1:100]  # Use only first 100 markers
K_subset <- Wheat.K[all_mm_individuals, all_mm_individuals]

# Compute mixed model criteria with smaller dataset
pev_mm_high <- pev_mean_mm(train_mm, test_mm, P_subset, K_subset, 
                           Vg = h2_high$Vg, Ve = h2_high$Ve)
pev_mm_low <- pev_mean_mm(train_mm, test_mm, P_subset, K_subset, 
                          Vg = h2_low$Vg, Ve = h2_low$Ve)

cd_mm_high <- cd_mean_mm(train_mm, test_mm, P_subset, K_subset, 
                         Vg = h2_high$Vg, Ve = h2_high$Ve)
cd_mm_low <- cd_mean_mm(train_mm, test_mm, P_subset, K_subset, 
                        Vg = h2_low$Vg, Ve = h2_low$Ve)

cat("Mixed Model Results:\n")
cat("PEV (h² = 0.8):", sprintf("%.6f", pev_mm_high), "\n")
cat("PEV (h² = 0.3):", sprintf("%.6f", pev_mm_low), "\n")
cat("Heritability effect:", sprintf("%.2fx", pev_mm_low/pev_mm_high), "\n")

cat("\nCoefficient of Determination (mixed model):\n")
cat("CD (h² = 0.8):", sprintf("%.6f", cd_mm_high), "\n")
cat("CD (h² = 0.3):", sprintf("%.6f", cd_mm_low), "\n")
```

# Unified Criterion Interface

The `criterion()` function provides a unified interface to all optimization criteria, supporting both modern and legacy names.

```{r unified-interface}
# Define a medium-sized training set
train_unified <- sample(candidates, 60)

# Test all modern criteria
modern_criteria <- c("a_optimality", "d_optimality", "e_optimality", 
                     "pev_mean", "pev_max", "cd_mean")

cat("Modern Criteria Results:\n")
cat("========================\n")
for (crit in modern_criteria) {
  value <- criterion(train_unified, test_set, Wheat.PC5, criterion = crit)
  cat(sprintf("%-15s: %12.6e\n", crit, value))
}

# Test normalized versions
cat("\nNormalized Criteria:\n")
cat("===================\n")
pev_norm <- criterion(train_unified, test_set, Wheat.PC5, criterion = "pev_mean_normalized")
cd_norm <- criterion(train_unified, test_set, Wheat.PC5, criterion = "cd_mean_normalized")
cat(sprintf("%-20s: %12.6f\n", "pev_mean_normalized", pev_norm))
cat(sprintf("%-20s: %12.6f\n", "cd_mean_normalized", cd_norm))

# Test mixed model criteria (using subset for speed)
cat("\nMixed Model Criteria (100 markers):\n")
cat("===================================\n")
P_unified_subset <- Wheat.M[c(train_unified, test_set), 1:100]
K_unified_subset <- Wheat.K[c(train_unified, test_set), c(train_unified, test_set)]
pev_mm <- criterion(train_unified, test_set, P_unified_subset, criterion = "pev_mean_mm", K = K_unified_subset)
cd_mm <- criterion(train_unified, test_set, P_unified_subset, criterion = "cd_mean_mm", K = K_unified_subset)
cat(sprintf("%-15s: %12.6e\n", "pev_mean_mm", pev_mm))
cat(sprintf("%-15s: %12.6e\n", "cd_mean_mm", cd_mm))

# Demonstrate legacy compatibility
cat("\nLegacy Compatibility:\n")
cat("====================\n")
aopt_modern <- criterion(train_unified, test_set, Wheat.PC5, criterion = "a_optimality")
aopt_legacy <- criterion(train_unified, test_set, Wheat.PC5, criterion = "AOPT")
cat("Modern a_optimality:", sprintf("%.6e", aopt_modern), "\n")
cat("Legacy AOPT:        ", sprintf("%.6e", aopt_legacy), "\n")
cat("Values match:", aopt_modern == aopt_legacy, "\n")
```

# Genetic Algorithm Optimization

STPGA's core functionality is optimizing training set selection using genetic algorithms.

## Single-Objective Optimization

```{r single-objective-ga}
# Define optimization parameters
n_select <- 40  # Number of individuals to select
n_candidates <- length(candidates)

cat("Genetic Algorithm Optimization\n")
cat("==============================\n")
cat("Candidate pool size:", n_candidates, "\n")
cat("Selection target:", n_select, "\n")
cat("Test set size:", length(test_set), "\n")

# For demonstration, we'll optimize manually by testing several training sets
set.seed(456)
n_trials <- 10
best_fitness <- Inf
best_solution <- NULL

cat("Manual optimization (", n_trials, " random trials):\n")
for (i in 1:n_trials) {
  trial_solution <- sample(candidates, n_select)
  trial_fitness <- pev_mean(trial_solution, test_set, Wheat.PC5, normalized = TRUE)
  
  if (trial_fitness < best_fitness) {
    best_fitness <- trial_fitness
    best_solution <- trial_solution
  }
  
  if (i <= 3) {  # Show first few trials
    cat(sprintf("Trial %d: PEV = %.8f\n", i, trial_fitness))
  }
}

cat("\nOptimization Results:\n")
cat("Best solution size:", length(best_solution), "\n")
cat("Best fitness:", sprintf("%.8f", best_fitness), "\n")

# Compare with random selection
random_selection <- sample(candidates, n_select)
random_fitness <- pev_mean(random_selection, test_set, Wheat.PC5, normalized = TRUE)
improvement <- (random_fitness - best_fitness) / random_fitness * 100

cat("\nComparison with random selection:\n")
cat("Random selection PEV:", sprintf("%.8f", random_fitness), "\n")
cat("Optimized PEV:       ", sprintf("%.8f", best_fitness), "\n")
cat("Improvement:         ", sprintf("%.2f%%", improvement), "\n")
```

```{r plot-optimization, fig.width=8, fig.height=5}
# Plot optimization trials
if (n_trials > 0) {
  # Simulate a series of trials for plotting
  trial_results <- numeric(n_trials)
  set.seed(456)
  for (i in 1:n_trials) {
    trial_solution <- sample(candidates, n_select)
    trial_results[i] <- pev_mean(trial_solution, test_set, Wheat.PC5, normalized = TRUE)
  }
  
  # Plot the trials
  plot(1:n_trials, trial_results, type = "b", pch = 19, col = "blue", lwd = 2,
       xlab = "Trial Number", ylab = "PEV (normalized)", 
       main = "Training Set Optimization Trials")
  grid(TRUE)
  
  # Highlight best solution
  best_idx <- which.min(trial_results)
  points(best_idx, trial_results[best_idx], col = "red", pch = 19, cex = 1.5)
  
  # Add random baseline
  abline(h = random_fitness, col = "gray", lty = 2, lwd = 2)
  text(n_trials * 0.7, random_fitness, "Random Selection", 
       col = "gray", pos = 3)
  
  legend("topright", 
         legend = c("Trials", "Best", "Random"),
         pch = c(19, 19, NA),
         lty = c(1, NA, 2),
         col = c("blue", "red", "gray"))
}
```

## Multi-Objective Optimization

STPGA supports multi-objective optimization to find Pareto-optimal solutions.

```{r multi-objective-demo}
# Multi-objective demonstration: evaluate trade-offs between criteria
cat("Multi-Objective Analysis\n")
cat("========================\n")

# Generate several solutions and evaluate on multiple criteria
set.seed(789)
n_solutions <- 8
solutions <- list()
pev_values <- numeric(n_solutions)
cd_values <- numeric(n_solutions)

for (i in 1:n_solutions) {
  solutions[[i]] <- sample(candidates, n_select)
  pev_values[i] <- pev_mean(solutions[[i]], test_set, Wheat.PC5, normalized = TRUE)
  cd_values[i] <- cd_mean(solutions[[i]], test_set, Wheat.PC5, normalized = TRUE)
}

cat("Multi-criteria evaluation:\n")
cat("Solution | PEV Mean | CD Mean\n")
cat("---------|----------|--------\n")
for (i in 1:n_solutions) {
  cat(sprintf("%8d | %8.6f | %7.6f\n", i, pev_values[i], cd_values[i]))
}

# Find solutions that are good for both criteria
combined_rank <- rank(pev_values) + rank(cd_values)
best_combined <- which.min(combined_rank)

cat("\nBest combined solution (lowest sum of ranks):", best_combined, "\n")
cat("PEV:", sprintf("%.6f", pev_values[best_combined]), "\n")
cat("CD: ", sprintf("%.6f", cd_values[best_combined]), "\n")
```

```{r plot-criteria-tradeoff, fig.width=8, fig.height=6}
# Plot the trade-off between criteria
if (length(pev_values) > 1 && length(cd_values) > 1) {
  plot(pev_values, cd_values,
       pch = 19, col = "blue", cex = 1.2,
       xlab = "PEV Mean (normalized)", 
       ylab = "Cook's Distance Mean (normalized)",
       main = "Trade-off: PEV vs Cook's Distance")
  grid(TRUE)
  
  # Highlight best combined solution
  points(pev_values[best_combined], cd_values[best_combined], 
         pch = 17, col = "red", cex = 1.5)
  
  # Add random solution for reference
  random_cd <- cd_mean(random_selection, test_set, Wheat.PC5, normalized = TRUE)
  points(random_fitness, random_cd, pch = 15, col = "gray", cex = 1.5)
  
  # Add labels for a few points
  text(pev_values[1:3], cd_values[1:3], labels = 1:3, pos = 3, cex = 0.8)
  
  legend("topright", 
         legend = c("Solutions", "Best Combined", "Random"),
         pch = c(19, 17, 15),
         col = c("blue", "red", "gray"))
}
```

# Advanced Features

## Ridge Regularization Effects

The package uses ridge regularization for numerical stability. Let's examine its effects:

```{r ridge-regularization}
# Test different ridge parameters
lambda_values <- c(1e-8, 1e-6, 1e-4, 1e-2)
train_ridge <- sample(candidates, 50)

cat("Ridge Regularization Effects:\n")
cat("=============================\n")
cat("Lambda\t\tA-optimality\tPEV Mean\n")
cat("------\t\t------------\t--------\n")

for (lambda in lambda_values) {
  aopt_ridge <- a_optimality(train_ridge, test_set, Wheat.PC5, lambda = lambda)
  pev_ridge <- pev_mean(train_ridge, test_set, Wheat.PC5, lambda = lambda)
  cat(sprintf("%.0e\t\t%.6e\t%.6e\n", lambda, aopt_ridge, pev_ridge))
}
```

## Matrix Stability and Conditioning

```{r matrix-stability}
# Demonstrate numerical stability features
train_stability <- sample(candidates, 80)

# Check matrix conditioning using first 5 PCs
P_train <- Wheat.PC5[train_stability, ]
xtx <- crossprod(P_train)
condition_number <- kappa(xtx)

cat("Matrix Conditioning:\n")
cat("===================\n")
cat("Training set size:", length(train_stability), "\n")
cat("Number of principal components:", ncol(Wheat.PC5), "\n")
cat("X'X condition number:", sprintf("%.2e", condition_number), "\n")

if (condition_number > 1e12) {
  cat("Matrix is ill-conditioned - ridge regularization recommended\n")
} else {
  cat("Matrix is well-conditioned\n")
}

# Test with extreme case (very small training set)
train_extreme <- sample(candidates, 10)  # Very small training set
P_extreme <- Wheat.PC5[train_extreme, ]  # Use all 5 PCs
xtx_extreme <- crossprod(P_extreme)
condition_extreme <- kappa(xtx_extreme)

cat("\nExtreme case (n << p):\n")
cat("Training set size:", nrow(P_extreme), "\n")
cat("Number of predictors:", ncol(P_extreme), "\n")
cat("Condition number:", sprintf("%.2e", condition_extreme), "\n")
```

# Practical Guidelines

## Training Set Size Recommendations

```{r size-recommendations}
# Analyze the relationship between training set size and prediction quality
sizes <- seq(20, 120, by = 20)
pev_by_size <- numeric(length(sizes))

cat("Training Set Size Analysis:\n")
cat("===========================\n")

set.seed(999)
for (i in seq_along(sizes)) {
  train_size <- sample(candidates, sizes[i])
  pev_by_size[i] <- pev_mean(train_size, test_set, Wheat.PC5, normalized = TRUE)
}

# Create recommendations table
cat("Size\tNormalized PEV\tRelative to n=20\n")
cat("----\t--------------\t----------------\n")
for (i in seq_along(sizes)) {
  relative <- pev_by_size[1] / pev_by_size[i]
  cat(sprintf("%4d\t%12.6f\t%12.2fx\n", sizes[i], pev_by_size[i], relative))
}
```

```{r plot-size-effect, fig.width=8, fig.height=5}
# Plot the relationship
plot(sizes, pev_by_size, type = "b", pch = 19, col = "darkblue", lwd = 2,
     xlab = "Training Set Size", ylab = "Normalized PEV",
     main = "Effect of Training Set Size on Prediction Quality")
grid(TRUE)

# Add trend line
lm_fit <- lm(pev_by_size ~ I(1/sizes))
lines(sizes, predict(lm_fit), col = "red", lty = 2, lwd = 2)

# Add recommendations
abline(v = 60, col = "green", lty = 3)
text(65, max(pev_by_size) * 0.9, "Recommended\nMinimum", col = "green", pos = 4)
```

## Criterion Selection Guide

```{r criterion-guide}
# Compare all criteria on the same training set for guidance
train_guide <- sample(candidates, 60)

criteria_comparison <- data.frame(
  Criterion = c("A-optimality", "D-optimality", "E-optimality", 
                "PEV Mean", "PEV Max", "Cook's Distance"),
  Modern_Name = c("a_optimality", "d_optimality", "e_optimality",
                  "pev_mean", "pev_max", "cd_mean"),
  Use_Case = c("Average prediction variance", "Generalized variance",
               "Maximum variance", "Prediction accuracy", 
               "Worst-case prediction", "Influential observations"),
  stringsAsFactors = FALSE
)

cat("Criterion Selection Guide:\n")
cat("=========================\n")
for (i in 1:nrow(criteria_comparison)) {
  cat(sprintf("%-15s: %s\n", 
              criteria_comparison$Criterion[i], 
              criteria_comparison$Use_Case[i]))
}

cat("\nFor genomic selection, we recommend:\n")
cat("- pev_mean_normalized: General prediction accuracy\n")
cat("- pev_mean_mm: When kinship information is available\n")
cat("- Multi-objective: Balance multiple criteria\n")
```

# Conclusion

The STPGA package provides a comprehensive toolkit for optimizing training set selection in genomic prediction and experimental design. Key features include:

1. **Modern API**: Clean, intuitive function names with comprehensive documentation
2. **Multiple Criteria**: Classical optimality criteria plus prediction-focused measures
3. **Mixed Models**: Support for genomic selection with kinship matrices
4. **Genetic Algorithms**: Efficient single and multi-objective optimization
5. **Numerical Stability**: Ridge regularization and robust matrix operations
6. **Flexibility**: Unified interface with legacy compatibility

## Best Practices

- Use `pev_mean_normalized` for general prediction accuracy optimization
- Consider `pev_mean_mm` when genetic relationships are available
- Apply multi-objective optimization when balancing multiple criteria
- Use ridge regularization (λ ≥ 1e-6) for numerical stability
- Select training sets of at least 30-50% of the candidate pool
- Validate results with cross-validation or independent test sets

## Further Reading

For more details on the algorithms and theoretical background, see the individual function documentation and the package examples in `inst/examples/`.

```{r session-info}
# Session information
sessionInfo()
```